---
title: "Spatial Modeling"
subtitle: "Enhanced Geographic Analysis Using Official NYC Boundaries"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
---

<style>
.hero-spatial {
  background: linear-gradient(135deg, #667eea 0%, #764ba2 100%);
  color: white;
  padding: 40px;
  border-radius: 15px;
  margin: 30px 0;
  text-align: center;
}

.map-section {
  background: white;
  padding: 25px;
  border-radius: 15px;
  box-shadow: 0 4px 20px rgba(0,0,0,0.08);
  margin: 25px 0;
}

.spatial-insight {
  background: linear-gradient(to right, #f0f9ff, #e0f2fe);
  border-left: 5px solid #2563eb;
  padding: 20px;
  margin: 20px 0;
  border-radius: 8px;
}

.warning-box {
  background: #fef3c7;
  border-left: 5px solid #f59e0b;
  padding: 15px;
  margin: 15px 0;
  border-radius: 5px;
}

.leaflet-container {
  height: 600px !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8
)

# Load packages
library(tidyverse)
library(sf)
library(leaflet)
library(leaflet.extras)
library(viridis)
library(plotly)
library(RColorBrewer)
library(DT)
library(tigris)  # For downloading Census shapefiles
```

<div class="hero-spatial">
<h1>🗽 NYC Spatial Analysis with Real Boundaries</h1>
<p style="font-size: 1.2rem;">Using Official NYC Borough Shapefiles for Accurate Geographic Visualization</p>
</div>

---

```{r, include=FALSE}
# Load your PM2.5 and asthma data
pm25_asthma <- read_csv("data/merged_pm25_asthma.csv") %>%
  mutate(
    borough = factor(borough, 
                     levels = c("Manhattan", "Bronx", "Brooklyn", 
                                "Queens", "Staten Island"))
  )
```

## Use tigris Package (Automated) to download Census boundaries

```{r message=FALSE, warning=FALSE, results='hide'}
# Set tigris options
options(tigris_use_cache = TRUE)

# Get New York State counties
ny_counties <- counties(state = "NY", cb = TRUE, year = 2023)

# NYC is made up of these 5 counties:
nyc_county_names <- c("New York", "Kings", "Queens", "Bronx", "Richmond")

# Filter for NYC counties
nyc_boroughs_tiger <- ny_counties %>%
  filter(NAME %in% nyc_county_names)

# Map county names to borough names
nyc_boroughs <- nyc_boroughs_tiger %>%
  mutate(
    borough = case_when(
      NAME == "New York" ~ "Manhattan",
      NAME == "Kings" ~ "Brooklyn",
      NAME == "Queens" ~ "Queens",
      NAME == "Bronx" ~ "Bronx",
      NAME == "Richmond" ~ "Staten Island"
    ),
    borough = factor(borough, 
                     levels = c("Manhattan", "Bronx", "Brooklyn", 
                                "Queens", "Staten Island"))
  ) %>%
  select(borough, NAME, geometry)

# Check the shapefile
glimpse(nyc_boroughs)
```

```{r show-shapefile-structure}
# Display shapefile information
cat("CRS (Coordinate Reference System):\n")
st_crs(nyc_boroughs)

cat("\nBoroughs in shapefile:\n")
nyc_boroughs %>% 
  st_drop_geometry() %>%
  select(borough) %>%
  knitr::kable()
```

## Merge with Data

```{r merge-spatial-data}
# For 2023 data
spatial_data_2023 <- pm25_asthma %>%
  filter(year == 2023) %>%
  right_join(nyc_boroughs, by = "borough") %>%
  st_as_sf()

# For all years
spatial_data_full <- pm25_asthma %>%
  right_join(nyc_boroughs, by = "borough") %>%
  st_as_sf()

# Verify the merge
cat("Data merged successfully\n")
cat("Number of features:", nrow(spatial_data_2023), "\n")
cat("Columns:", paste(names(spatial_data_2023), collapse = ", "))
```

---

# Interactive Choropleth Maps {.tabset .tabset-pills}

## PM2.5 Choropleth (2023)

<div class="map-section">

```{r choropleth-pm25-2023, fig.height=7}
# Create color palette
pal_pm25 <- colorNumeric(
  palette = "YlOrRd",
  domain = spatial_data_2023$pm25_annual,
  na.color = "transparent"
)

# Create choropleth map
leaflet(spatial_data_2023) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_pm25(pm25_annual),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0(
      "<strong>", borough, "</strong><br/>",
      "PM2.5: ", round(pm25_annual, 1), " μg/m³<br/>",
      "Summer: ", round(pm25_summer, 1), " μg/m³<br/>",
      "Winter: ", round(pm25_winter, 1), " μg/m³"
    ) %>% lapply(htmltools::HTML),
    popup = ~paste0(
      "<div style='font-size: 14px;'>",
      "<h4 style='color: #2563eb; margin-bottom: 10px;'>", borough, "</h4>",
      "<strong>Annual PM2.5:</strong> ", round(pm25_annual, 1), " μg/m³<br/>",
      "<strong>Summer PM2.5:</strong> ", round(pm25_summer, 1), " μg/m³<br/>",
      "<strong>Winter PM2.5:</strong> ", round(pm25_winter, 1), " μg/m³<br/>",
      "<strong>Asthma Rate:</strong> ", round(asthma_rate, 1), " per 10,000",
      "</div>"
    ) %>% lapply(htmltools::HTML),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "13px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_pm25,
    values = ~pm25_annual,
    title = "PM2.5 (μg/m³)",
    opacity = 0.7
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  setView(lng = -73.935242, lat = 40.730610, zoom = 10)
```

</div>

## Asthma Rate Choropleth (2023)

<div class="map-section">

```{r choropleth-asthma-2023, fig.height=7}
# Create color palette
pal_asthma <- colorNumeric(
  palette = "Reds",
  domain = spatial_data_2023$asthma_rate,
  na.color = "transparent"
)

leaflet(spatial_data_2023) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_asthma(asthma_rate),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0(
      "<strong>", borough, "</strong><br/>",
      "Asthma Rate: ", round(asthma_rate, 1), " per 10,000<br/>",
      "PM2.5: ", round(pm25_annual, 1), " μg/m³"
    ) %>% lapply(htmltools::HTML),
    popup = ~paste0(
      "<div style='font-size: 14px;'>",
      "<h4 style='color: #dc2626; margin-bottom: 10px;'>", borough, "</h4>",
      "<strong>Asthma ED Rate:</strong> ", round(asthma_rate, 1), " per 10,000<br/>",
      "<strong>PM2.5 Level:</strong> ", round(pm25_annual, 1), " μg/m³<br/>",
      "<strong>Ratio:</strong> ", round(asthma_rate / pm25_annual, 1), " asthma per PM2.5",
      "</div>"
    ) %>% lapply(htmltools::HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_asthma,
    values = ~asthma_rate,
    title = "Asthma Rate<br/>(per 10,000)",
    opacity = 0.7
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  setView(lng = -73.935242, lat = 40.730610, zoom = 10)
```

</div>

<div class="spatial-insight">
<h4>🏥 Geographic Health Disparity</h4>
The choropleth clearly shows **The Bronx** in dark red, visualizing the disproportionate asthma burden. The actual boundary visualization makes the environmental justice issue more tangible and impactful.
</div>

## Side-by-Side Comparison

<div class="map-section">

```{r side-by-side-choropleth, fig.height=7}
# Create synchronized maps
leaflet(spatial_data_2023) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # PM2.5 polygons
  addPolygons(
    fillColor = ~pal_pm25(pm25_annual),
    weight = 2,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    group = "PM2.5",
    label = ~paste0(borough, ": ", round(pm25_annual, 1), " μg/m³") %>% 
      lapply(htmltools::HTML)
  ) %>%
  # Asthma polygons
  addPolygons(
    fillColor = ~pal_asthma(asthma_rate),
    weight = 2,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    group = "Asthma Rate",
    label = ~paste0(borough, ": ", round(asthma_rate, 1), " per 10,000") %>% 
      lapply(htmltools::HTML)
  ) %>%
  addLayersControl(
    baseGroups = c("PM2.5", "Asthma Rate"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("#fee5d9", "#a50f15"),
    labels = c("Lower", "Higher"),
    title = "Levels",
    opacity = 0.7
  ) %>%
  setView(lng = -73.935242, lat = 40.730610, zoom = 10)
```

</div>

---

# Static Choropleth Maps

## Using ggplot2 + sf

```{r ggplot-choropleth, fig.height=10, fig.width=12}
library(patchwork)

# PM2.5 map
p1 <- ggplot(spatial_data_2023) +
  geom_sf(aes(fill = pm25_annual), color = "white", size = 0.5) +
  geom_sf_text(aes(label = borough), size = 3, fontface = "bold") +
  scale_fill_viridis_c(
    option = "plasma",
    name = "PM2.5\n(μg/m³)",
    direction = -1
  ) +
  labs(
    title = "Annual PM2.5 Concentration by Borough (2023)",
    subtitle = "Darker colors indicate higher pollution levels"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

# Asthma map
p2 <- ggplot(spatial_data_2023) +
  geom_sf(aes(fill = asthma_rate), color = "white", size = 0.5) +
  geom_sf_text(aes(label = borough), size = 3, fontface = "bold") +
  scale_fill_viridis_c(
    option = "inferno",
    name = "Asthma Rate\n(per 10,000)",
    direction = -1
  ) +
  labs(
    title = "Asthma ED Visit Rate by Borough (2023)",
    subtitle = "The Bronx shows dramatically elevated rates"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

p1 / p2
```

---

# Multi-Year Evolution

## Faceted Choropleth Maps

```{r faceted-choropleth, fig.height=12, fig.width=14}
# Select key years
key_years <- c(2009, 2013, 2017, 2020, 2023)

spatial_data_years <- pm25_asthma %>%
  filter(year %in% key_years) %>%
  right_join(nyc_boroughs, by = "borough") %>%
  st_as_sf()

# PM2.5 evolution
ggplot(spatial_data_years) +
  geom_sf(aes(fill = pm25_annual), color = "white", size = 0.3) +
  geom_sf_text(aes(label = borough), size = 2, fontface = "bold") +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "PM2.5\n(μg/m³)",
    direction = -1,
    limits = c(5, 13)  # Fixed scale for comparison
  ) +
  labs(
    title = "PM2.5 Concentration Evolution Across NYC Boroughs",
    subtitle = "Borough boundaries with pollution levels | 2009-2023"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40"),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

```{r faceted-asthma-choropleth, fig.height=12, fig.width=14}
# Asthma evolution
ggplot(spatial_data_years) +
  geom_sf(aes(fill = asthma_rate), color = "white", size = 0.3) +
  geom_sf_text(aes(label = borough), size = 2, fontface = "bold") +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_viridis_c(
    option = "inferno",
    name = "Asthma Rate\n(per 10,000)",
    direction = -1,
    limits = c(20, 230)  # Fixed scale
  ) +
  labs(
    title = "Asthma ED Rate Evolution Across NYC Boroughs",
    subtitle = "Geographic health disparities over time | 2009-2023"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40"),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

---

# Spatial Statistics with Real Boundaries

## Calculate Area-Based Metrics

```{r area-calculations}
# Calculate area of each borough
borough_areas <- spatial_data_2023 %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,  # Convert to km²
    pm25_per_km2 = pm25_annual / area_km2,
    asthma_per_km2 = asthma_rate / area_km2
  ) %>%
  st_drop_geometry()

# Display table
datatable(
  borough_areas %>%
    select(borough, area_km2, pm25_annual, asthma_rate, pm25_per_km2, asthma_per_km2) %>%
    mutate(across(where(is.numeric), ~round(., 2))),
  caption = "Borough Statistics Including Area",
  options = list(pageLength = 5, dom = 't'),
  rownames = FALSE
)
```

## Spatial Neighbors Analysis

```{r spatial-neighbors}
# Create spatial weights based on queen contiguity
library(spdep)

# Get neighbors
neighbors <- poly2nb(spatial_data_2023, queen = TRUE)

# Create summary
neighbor_summary <- tibble(
  Borough = spatial_data_2023$borough,
  `Number of Neighbors` = card(neighbors),
  `Neighbors` = sapply(neighbors, function(x) {
    if (length(x) == 0) return("None (island)")
    paste(spatial_data_2023$borough[x], collapse = ", ")
  })
)

knitr::kable(
  neighbor_summary,
  caption = "Spatial Contiguity: Which Boroughs Share Boundaries?"
)
```

---

# Key Findings with Real Boundaries

<div class="map-section">

## Geographic Insights

1. **Manhattan's Density**: Despite being the smallest borough by area, shows highest PM2.5 concentration
2. **Bronx's Burden**: Medium-sized borough with disproportionately high asthma rates
3. **Brooklyn's Position**: Largest population, moderate pollution and asthma levels
4. **Queens' Extent**: Largest by area, relatively lower pollution and health impacts
5. **Staten Island**: Most isolated (no shared boundaries), lowest pollution

## Spatial Patterns

- **Contiguity matters**: Bronx borders Manhattan (high PM2.5) but has lower pollution
- **Size-health relationship**: Borough size ≠ health burden
- **Environmental justice**: Visible geographic concentration of health disparities

</div>
